Transfer of Learned Opponent Models in Zero Sum Games

### 2020-01-27 Report/draft_report_v1.tex
added naive tom and qls with transfer

In this paper, we are specifically interested in the way in which people build and use models of their opponent to facilitate learning transfer, when engaged in situations involving an interaction with strategic considerations. These situations arise frequently such as in negotiations, auctions, strategic planning and all other domains in which theory of mind abilities (Premack \& Woodruff, 1978) play a role in determining human behaviour.

@@

Early literature on learning transfer in games has mostly focused on measuring the proportion of people who play normatively optimal (Nash Equilibria) or salient actions (e.g Risk Dominance) in later games, having had experience with a similar game environment previously. For instance, Ho et al. (1998) measure transfer as the proportion of players who choose the Nash Equilibrium in later p-beauty contest games, after training on similar games. They find there is no evidence of immediate transfer (Nash equilibrium play in the first round of the new game) but positive structural learning transfer as shown by the faster convergence to equilibrium play by experienced vs non experienced players. Camerer \& Knez (2000) test learning transfer in players exposed to two games with multiple equilibria sequentially and explore the ability of players to coordinate their actions to choose a particular equilibrium in subsequent games having reached it in prior ones. They distinguish between games that are similar in a purely descriptive way, meaning similar choice labels, identity of players, format and number of action choices; and games that are similar in a strategic sense, meaning similar payoffs from combination of actions, identical equilibrium properties or significant social characteristics of payoffs such as possibility of punishment, need for fairness and cooperative vs competitive settings. They find that transfer of learning (successful coordination) occurs more readily in the presence of both descriptive and strategic similarity. If the games were only strategically similar, then the transfer was much weaker.

Juvina et al.(2014) made a similar distinction between what they deemed surface and deep similarities and find that both contribute to positive learning transfer. However, they show that surface similarity is not necessary for deep transfer and can either aid or block this type of transfer depending on whether it leads to congruent or incongruent actions in later games. In a series of experiments using economic signalling games, Cooper \& Kagel (2003, 2004, 2008) find that participants who have learned to play according to Nash Equilibrium in one game can transfer this to subsequent games, even though the actions consistent with Nash Equilibrium in later games are different. They show that this transfer is driven by the emergence of sophisticated players who are able to represent the strategic implications of their actions and reason about the consequences of changed opponent payoffs.

Most of these studies fail to offer a formal explanation of this transfer or a modelling framework that can explain the experimental observation of transfer between games and generalise it to extensive classes of games. A notable exception is the effort by Haruvy and Stahl (2012) to specify a model of learning where players learn abstract rules that they can generalise and transfer across dissimilar games, rather than action choices that can only be used within the same game. Participants played ten games, presented as 4x4 normal form (matrix payoffs). Their results suggest that subjects do transfer learning over descriptively similar but strategically dissimilar games and that this learning transfer is significant. They also showed that players learn abstract aspects of the game that are then transferred to new settings. Their rule-learning model, based on Stahl (1996), was able to capture participants' dynamic behavior and show that the propensity to select particular rules is perfectly transferred across games.

Still, one of the commonalities in studies of how humans adapt to computerised opponents is that they have mostly looked at the ability of players to detect and exploit action-based learning rules. The strategies implemented by the computer opponents had a style of play that was not ``human-like'' in the sense that humans are not very good at playing specific mixed strategies with any precision of at detecting patterns from long sequences of past play due to cognitive constraints. It is therefore important to have agents that ``play like humans'', and one way of achieving that is to embed theses agents with human-like theory of mind abilities based on limited steps of recursive reasoning. Simon (1972) explains that humans have limited cognitive capacities and as such cannot be expected to solve computationally intractable problems such as finding Nash equilibria. Instead, they will try to `satisfice' by choosing a strategy that is adequate in a simplified model of the environment, rather than an optimal one. This concept finds its natural application in `level-k' theory, first adopted by Stahl \& Wilson (1995). It posits that deviations from Nash equilibrium solutions in simple games are explained by the fact that humans have a heterogenous degree of strategic sophistication. At the bottom of the ladder, level-0 players are non-strategic and play either randomly or use a salient strategy in the game environment (Arad \& Rubinstein, 2012). Level-1 players are next up the ladder of strategic sophistication and will assume all their opponents belong to the level-0 category and as such will best respond to them given this assumption. Likewise, a level-2 player will choose actions that are the best response given the belief that all opponents are exactly one level below, and so on.

In this study, we propose to explore opponent modelling and its transfer with the use of computer agents possessing human-like theory of mind abilities with limited degrees of iterated reasoning. The agents will have a fixed strategy played stochastically and will either be level-1 or level-2 behavioral rule, mimicking human theory of mind abilities and the limited recursion depth they exhibit (Goodie et al., 2012).

Our choice of using computer oponents sinstead of matching groups of participants makes it easier to disentangle the process of learning about the opponent from that of learning about the game structure and payoffs. When playing against other human opponents,players are learning about the game and the strategy of the opponent simultaneously. Thus, it is harder to focus on an individual and how her strategies are changing and adapting to the opponent's play if we cannot experimentally control the behaviour of the opponent. The use of computer opponents to elicit learning behavior has been explored in the literature with encouraging results. For instance, Spiliopoulos (2013) made humans play constant sum games against 3 computer opponents programmed to take advantage of known patterns in human play such as imperfect randomization and heuristics use and found that human participants do adapt to the opponent they are facing. Shachat \& Swarthout (2004) made human participants face computer opponents playing various mixed strategies in a zero-sum asymmetric matching pennies game. They found that the players changed their strategies towards exploiting the deviations from the Mixed Strategy Nash Equilibrium (MSNE), and that this exploitation was very likely if the deviation from the MSNE play was high.

@@

We extend this very simple model by adding a state space that consists of last round human and agent play. This is akin to using a Q-learning algorithm ({\textbf{???}}). The update rule becomes:

@@

Next, we use a self-tuning Experience Weighted attraction model ({\textbf{???}}). EWA models particularity is that thet nest two seemingly different approaches, namely reinforcement learning and belief learning. Belief based models are based on the assumption that players keep track of the frequency of past plays and best respond to that. In contrast, reinforcement learning does not take into account beliefs about other players, but is such that an action followed by a positive reward is more likely to be repeated than an action followed by a negative reward. The self-tuning EWA model has been shown to perform better than both these nested models in multiple repeated gmes and has the advantage of having only one free parameter, the inverse temperature in the softmax choice function.

@@

Here, I(x,y) is the indicator function equla to 1 if \$ x = y\$ and 0 otherwise. A simple way to think about this update rule is that attractions are multipled by a parameter that represents experince (\(N(t-1)\)) which is itself decaying by a wight \(\phi\). The result is added to either the payoff received (when the indicator function is 0), or to \(\delta\) times the foregone payoff (when indicator function is 1). We can see that setting \(\delta = 0\) leads to reinforcement of past actions, while positive and high delta parameters make the update rule take into account foregone payoffs, which is similar to weighted fictitous play ({\textbf{???}}).
While the assumption in expanding the update rule above is that \(\phi\) and \(\delta\) are free parameters ({\textbf{???}}), the self-tuning aspect of the model comes from the fact that these are now self-tuned using the formulas expanded in ({\textbf{???}}).

@@

Next we compared the performance of players whose actions are consistent with each of our hypothesized models. Figure~\ref{fig:exp1-cumScores} shows the average cumulative performance of players across games, for participants grouped by which model best fits their behavior in experiment 1. We can see that participants whose actions are most consistent with learning an opponent model in a Bayesian way had the best overall performance (both with and witout transfer), followed by Q-learning conditional on last round play. EWA, QL and random players had, understandably the lowest performance.

@@

{\centering \includegraphics{draft_report_v1_files/figure-latex/exp1-comp-model-1}

@@

\caption{Histogram of best fitting computational models for experiment 1}\label{fig:exp1-comp-model}
\end{figure}

@@

Plotting cumulative scores by best model for experiment 2, we see very similar results looking at Figure~\ref{fig:exp2-cumScores}, in that participants whose behavior was best fit by a bayesian model of learning the oppponent strategy had the highest cumulative performance, followed by theose relying on reinforcement learning with a state space of last round play, and as expectedm random play as well as EWA were at the bottom of cumulative performance.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-lake2017building}{}%
Lake, B. M., Ullman, T. D., Tenenbaum, J. B., \& Gershman, S. J. (2017). Building machines that learn and think like people. \emph{Behavioral and Brain Sciences}, \emph{40}.

>>>>

One way to build abstract representations of the environment when the task involves interactions with others is to build a model of the opponent strategy that may inform what actions they are likely to take next. Once we learn about the opponent's strategy, we can use this knowledge to inform how to best act in novel situations involving a previously encountered opponent. This may lead to very efficient generalization of knowledge, even to situations that are dissimilar to the history of interaction, assuming what we have learned about the opponent is an abstract representation that is not too dependent on the environment of the initial interaction. There is evidence that people learn models of the opponents when they play repeated economic games (Stahl \& Wilson, 1995), engage in bilateral negotiations (Baarslag, Hendrikx, Hindriks, \& Jonker, 2016), or simply try to exploit a non random player in chance games such as Rock-Paper-Scissors (Weerd, Verbrugge, \& Verheij, 2012). In this paper, we are specifically interested in the way in which people build and use models of their opponent to facilitate learning transfer, when engaged in situations involving an interaction with strategic considerations. These situations arise frequently such as in negotiations, auctions, strategic planning and all other domains in which theory of mind abilities (Woodruff, Sciences, Brain, \& 1978, n.d.) play a role in determining human behaviour.

@@

Early literature on learning transfer in games has mostly focused on measuring the proportion of people who play normatively optimal (Nash Equilibria) or salient actions (e.g Risk Dominance) in later games, having had experience with a similar game environment previously. For instance, Ho, Camerer, and Weigelt (1998) measure transfer as the proportion of players who choose the Nash Equilibrium in later p-beauty contest games, after training on similar games. They find there is no evidence of immediate transfer (Nash equilibrium play in the first round of the new game) but positive structural learning transfer as shown by the faster convergence to equilibrium play by experienced vs non experienced players. Knez and Camerer (2000) test learning transfer in players exposed to two games with multiple equilibria sequentially and explore the ability of players to coordinate their actions to choose a particular equilibrium in subsequent games having reached it in prior ones. They distinguish between games that are similar in a purely descriptive way, meaning similar choice labels, identity of players, format and number of action choices; and games that are similar in a strategic sense, meaning similar payoffs from combination of actions, identical equilibrium properties or significant social characteristics of payoffs such as possibility of punishment, need for fairness and cooperative vs competitive settings. They find that transfer of learning (successful coordination) occurs more readily in the presence of both descriptive and strategic similarity. If the games were only strategically similar, then the transfer was much weaker.

Juvina, Saleem, Martin, Gonzalez, and Lebiere (2013) made a similar distinction between what they deemed surface and deep similarities and find that both contribute to positive learning transfer. However, they show that surface similarity is not necessary for deep transfer and can either aid or block this type of transfer depending on whether it leads to congruent or incongruent actions in later games. In a series of experiments using economic signalling games (Cooper \& Kagel, 2003, 2008) the researchers found that participants who have learned to play according to Nash Equilibrium in one game can transfer this to subsequent games, even though the actions consistent with Nash Equilibrium in later games are different. They show that this transfer is driven by the emergence of sophisticated players who are able to represent the strategic implications of their actions and reason about the consequences of changed opponent payoffs.

Most of these studies fail to offer a formal explanation of this transfer or a modelling framework that can explain the experimental observation of transfer between games and generalise it to extensive classes of games. A notable exception is the effort by Haruvy and Stahl (2012) to specify a model of learning where players learn abstract rules that they can generalise and transfer across dissimilar games, rather than action choices that can only be used within the same game. Participants played ten games, presented as 4x4 normal form (matrix payoffs). Their results suggest that subjects do transfer learning over descriptively similar but strategically dissimilar games and that this learning transfer is significant. They also showed that players learn abstract aspects of the game that are then transferred to new settings. Their rule-learning model, based on Stahl and Wilson (1995), was able to capture participants' dynamic behavior and show that the propensity to select particular rules is perfectly transferred across games.

Still, one of the commonalities in studies of how humans adapt to computerised opponents is that they have mostly looked at the ability of players to detect and exploit action-based learning rules. The strategies implemented by the computer opponents had a style of play that was not ``human-like'' in the sense that humans are not very good at playing specific mixed strategies with any precision of at detecting patterns from long sequences of past play due to cognitive constraints. It is therefore important to have agents that ``play like humans'', and one way of achieving that is to embed theses agents with human-like theory of mind abilities based on limited steps of recursive reasoning. Simon (1972) explains that humans have limited cognitive capacities and as such cannot be expected to solve computationally intractable problems such as finding Nash equilibria. Instead, they will try to `satisfice' by choosing a strategy that is adequate in a simplified model of the environment, rather than an optimal one. This concept finds its natural application in `level-k' theory, first adopted by Stahl and Wilson (1995). It posits that deviations from Nash equilibrium solutions in simple games are explained by the fact that humans have a heterogenous degree of strategic sophistication. At the bottom of the ladder, level-0 players are non-strategic and play either randomly or use a salient strategy in the game environment Arad and Rubinstein (2012). Level-1 players are next up the ladder of strategic sophistication and will assume all their opponents belong to the level-0 category and as such will best respond to them given this assumption. Likewise, a level-2 player will choose actions that are the best response given the belief that all opponents are exactly one level below, and so on.

In this study, we propose to explore opponent modelling and its transfer with the use of computer agents possessing human-like theory of mind abilities with limited degrees of iterated reasoning. The agents will have a fixed strategy played stochastically and will either be level-1 or level-2 behavioral rule, mimicking human theory of mind abilities and the limited recursion depth they exhibit Goodie, Doshi, and Young (2012).

Our choice of using computer oponents sinstead of matching groups of participants makes it easier to disentangle the process of learning about the opponent from that of learning about the game structure and payoffs. When playing against other human opponents, players are learning about the game and the strategy of the opponent simultaneously. Thus, it is harder to focus on an individual and how her strategies are changing and adapting to the opponent's play if we cannot experimentally control the behaviour of the opponent. The use of computer opponents to elicit learning behavior has been explored in the literature with encouraging results. For instance, Spiliopoulos (2013) made humans play constant sum games against 3 computer opponents programmed to take advantage of known patterns in human play such as imperfect randomization and heuristics use and found that human participants do adapt to the opponent they are facing. Shachat and Swarthout (2004) made human participants face computer opponents playing various mixed strategies in a zero-sum asymmetric matching pennies game. They found that the players changed their strategies towards exploiting the deviations from the Mixed Strategy Nash Equilibrium (MSNE), and that this exploitation was very likely if the deviation from the MSNE play was high.

@@

We extend this very simple model by adding a state space that consists of last round human and agent play. This is akin to using a Q-learning algorithm (Watkins \& Dayan, 1992). The update rule becomes:

@@

Next, we use a self-tuning Experience Weighted attraction model (Ho, Camerer, \& Chong, 2004). EWA models particularity is that thet nest two seemingly different approaches, namely reinforcement learning and belief learning. Belief based models are based on the assumption that players keep track of the frequency of past plays and best respond to that. In contrast, reinforcement learning does not take into account beliefs about other players, but is such that an action followed by a positive reward is more likely to be repeated than an action followed by a negative reward. The self-tuning EWA model has been shown to perform better than both these nested models in multiple repeated gmes and has the advantage of having only one free parameter, the inverse temperature in the softmax choice function.

@@

Here, I(x,y) is the indicator function equla to 1 if \$ x = y\$ and 0 otherwise. A simple way to think about this update rule is that attractions are multipled by a parameter that represents experince (\(N(t-1)\)) which is itself decaying by a wight \(\phi\). The result is added to either the payoff received (when the indicator function is 0), or to \(\delta\) times the foregone payoff (when indicator function is 1). We can see that setting \(\delta = 0\) leads to reinforcement of past actions, while positive and high delta parameters make the update rule take into account foregone payoffs, which is similar to weighted fictitous play (Cheung \& Friedman, 1994).
While the assumption in expanding the update rule above is that \(\phi\) and \(\delta\) are free parameters (Camerer, Ho, \& Others, 1997), the self-tuning aspect of the model comes from the fact that these are now self-tuned using the formulas expanded in (Ho et al., 2004).

@@

Next we compared the performance of players whose actions are consistent with each of our hypothesized models. Figure~\ref{fig:exp1-cumScores} shows the average cumulative performance of players across games, for participants grouped by which model best fits their behavior in experiment 1. We can see that participants whose actions are most consistent with learning a ToM opponent model in a Bayesian way had the best overall performance (both with and witout transfer), followed by Q-learning conditional on last round play. EWA, QL and random players had, understandably the lowest performance.

@@

{\centering \includegraphics{draft_report_v1_files/figure-latex/exp1-cumScores-1}

@@

\caption{Average cumulative scores of participants in experiment 1 by best fitting model}\label{fig:exp1-cumScores}
\end{figure}

@@

Plotting cumulative scores by best model for experiment 2, we see very similar results looking at Figure~\ref{fig:exp2-cumScores}, in that participants whose behavior was best fit by a ToM model of learning the oppponent strategy had the highest cumulative performance, followed by theose relying on reinforcement learning with a state space of last round play, and as expected random play as well as EWA were at the bottom of cumulative performance.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-arad201211}{}%
Arad, A., \& Rubinstein, A. (2012). The 11-20 money request game: A level-k reasoning study. \emph{American Economic Review}, \emph{102}(7), 3561--3573.

\leavevmode\hypertarget{ref-baarslag2016learning}{}%
Baarslag, T., Hendrikx, M. J. C., Hindriks, K. V., \& Jonker, C. M. (2016). Learning about the opponent in automated bilateral negotiation: a comprehensive survey of opponent modeling techniques. \emph{Autonomous Agents and Multi-Agent Systems}, \emph{30}(5), 849--898.

\leavevmode\hypertarget{ref-camerer1997experience}{}%
Camerer, C., Ho, T.-H., \& Others. (1997). \emph{Experience-weighted attraction learning in games: A unifying approach}.

\leavevmode\hypertarget{ref-cheung1994learning}{}%
Cheung, Y.-W., \& Friedman, D. (1994). \emph{Learning in evolutionary games: some laboratory results}. University of California, Santa Cruz.

\leavevmode\hypertarget{ref-cooper2003lessons}{}%
Cooper, D. J., \& Kagel, J. H. (2003). Lessons learned: generalizing learning across games. \emph{American Economic Review}, \emph{93}(2), 202--207.

\leavevmode\hypertarget{ref-Cooper2008}{}%
Cooper, D. J., \& Kagel, J. H. (2008). Learning and transfer in signaling games. \emph{Economic Theory}, \emph{34}(3), 415--439. \url{https://doi.org/10.1007/s00199-006-0192-5}

\leavevmode\hypertarget{ref-goodie2012levels}{}%
Goodie, A. S., Doshi, P., \& Young, D. L. (2012). Levels of theory-of-mind reasoning in competitive games. \emph{Journal of Behavioral Decision Making}, \emph{25}(1), 95--108.

\leavevmode\hypertarget{ref-Haruvy2012}{}%
Haruvy, E., \& Stahl, D. O. (2012). Between-game rule learning in dissimilar symmetric normal-form games. \emph{Games and Economic Behavior}, \emph{74}(1), 208--221. \url{https://doi.org/10.1016/j.geb.2011.06.001}

\leavevmode\hypertarget{ref-ho2004economics}{}%
Ho, T. H., Camerer, C. F., \& Chong, J.-K. (2004). \emph{The economics of learning models: A self-tuning theory of learning in games}.

\leavevmode\hypertarget{ref-ho1998iterated}{}%
Ho, T.-H., Camerer, C., \& Weigelt, K. (1998). Iterated dominance and iterated best response in experimental" p-beauty contests". \emph{The American Economic Review}, \emph{88}(4), 947--969.

\leavevmode\hypertarget{ref-Juvina2013}{}%
Juvina, I., Saleem, M., Martin, J. M., Gonzalez, C., \& Lebiere, C. (2013). Reciprocal trust mediates deep transfer of learning between games of strategic interaction. \emph{Organizational Behavior and Human Decision Processes}, \emph{120}(2), 206--215. \url{https://doi.org/10.1016/j.obhdp.2012.09.004}

\leavevmode\hypertarget{ref-knez2000increasing}{}%
Knez, M., \& Camerer, C. (2000). Increasing cooperation in prisoner's dilemmas by establishing a precedent of efficiency in coordination games. \emph{Organizational Behavior and Human Decision Processes}, \emph{82}(2), 194--216.

\leavevmode\hypertarget{ref-Lake2017}{}%
Lake, B. M., Ullman, T. D., Tenenbaum, J. B., \& Gershman, S. J. (2017). Building machines that learn and think like people. \emph{Behavioral and Brain Sciences}, \emph{40}. \url{https://doi.org/10.1017/S0140525X16001837}

\leavevmode\hypertarget{ref-mertens1990repeated}{}%
Mertens, J.-F. (1990). Repeated games. In \emph{Game theory and applications} (pp. 77--130). Elsevier.

\leavevmode\hypertarget{ref-shachat2004we}{}%
Shachat, J., \& Swarthout, J. T. (2004). Do we detect and exploit mixed strategy play by opponents? \emph{Mathematical Methods of Operations Research}, \emph{59}(3), 359--373.

\leavevmode\hypertarget{ref-simon1972theories}{}%
Simon, H. A. (1972). Theories of bounded rationality. \emph{Decision and Organization}, \emph{1}(1), 161--176.

\leavevmode\hypertarget{ref-spiliopoulos2013strategic}{}%
Spiliopoulos, L. (2013). Strategic adaptation of humans playing computer algorithms in a repeated constant-sum game. \emph{Autonomous Agents and Multi-Agent Systems}, \emph{27}(1), 131--160.

\leavevmode\hypertarget{ref-stahl1995players}{}%
Stahl, D. O., \& Wilson, P. W. (1995). On players models of other players: Theory and experimental evidence. \emph{Games and Economic Behavior}, \emph{10}(1), 218--254.

\leavevmode\hypertarget{ref-watkins1992q}{}%
Watkins, C. J. C. H., \& Dayan, P. (1992). Q-learning. \emph{Machine Learning}, \emph{8}(3-4), 279--292.

\leavevmode\hypertarget{ref-de2012higher}{}%
Weerd, H. de, Verbrugge, R., \& Verheij, B. (2012). Higher-order social cognition in rock-paper-scissors: A simulation study. \emph{Proceedings of the 11th international conference on autonomous agents and multiagent systems-volume 3}, 1195--1196.

\leavevmode\hypertarget{ref-Woodruff}{}%
Woodruff, G., Sciences, D. P. B., Brain, \& 1978. (n.d.). \emph{Does the chimpanzee have a theory of mind}.
